Surrogate model for aligned-spin binary neutron star waveforms using gaussian process regression

### 2017-07-19 paper/paper.tex
More work on EOB section

%% ________________________________________________________
\section{Aligned-spin, dynamical tides EOB model}
Ever since its inception~\cite{Buonanno:1998gg}, the effective-one-body (EOB) approach to the general-relativistic 2-body problem has proven successful in modeling the dynamics and GW emission of binaries of compact objects. State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match numerical-relativity (NR) simulations of spinning, nonprecessing binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to almost extremal for equal-mass) binaries. The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}. This research program has been extended to accommodate matter effects~\cite{}, which become relevant XXX

References~\cite{Hinderer:2016eia,Steinhoff:2016rfi} built upon the aligned-spin BBH EOB model of Ref.~\cite{Taracchini:2013rva} and proposed a way to include the effect of dynamical tides. Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion --- be it a black hole or another neutron star. Because of binary motion, the forcing tidal field varies at the orbital frequency. Thus, in the late stages of the inspiral, the characteristic $f$-mode frequency of the neutron star can be dynamically approached, resulting in a resonant excitation of such mode. The net effect is an amplification of tidal effects as compared to the adiabatic limit, which assumes that the $f$-mode frequency is much larger than the frequency of the forcing tidal field. Dynamical tidal effects are implemented in the EOB model through a modification of the radial potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime. Here we adopt the tidally-augmented expression of $\Delta_u$ that is discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}, namely
\begin{equation}
\Delta_u = \Delta_u^{\textrm{pm}} + \Delta_u^{\textrm{DT}}\,,
\end{equation}
where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass term of the underlying BBH EOB model of Ref.~\cite{Bohe:2016gbl} (see Eq.~(2.2) therein) and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides. We consider the general case of a binary that may consist of two neutron stars; if either component of the binary is a black hole, then vanishing values of the tidal polarizabilities are employed in the formulas below. Let $A$ and $B$ be the body labels. Let $M_{A,B}$ be the mass of either body and $M=M_A+M_B$ be the total mass of the binary.  Explicitly, including quadrupolar and octupolar tidal effects, we have
\begin{widetext}
\begin{align}
 \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)X_A^4 X_B\,u^6  \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
 &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,,
\end{align}
\end{widetext}
where $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate, $\Lambda_{2,\textrm{dyn}}^{A,B}(u)$ are the dimensionless quadrupolar dynamical tidal polarizabilities, and $\Lambda_{3,\textrm{dyn}}^{A,B}(u)$ are the dimensionless octupolar dynamical tidal polarizabilities. Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies. In particular, the dimensionless dynamical polarizabilities read
\begin{equation}
\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})\,,
\end{equation}
where $k^{A,B}_{\ell}$ is the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ is the neutron star compactness, with $R_{A,B}$ the neutron star radius. Here $\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor given in Eq.~(11) of Ref.~\cite{Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole angular frequency, $\omega_{0\ell}^{A,B}$.

Tidal effects also enter the dissipative part of the model. In particular, the point-mass $\ell=2,3$ inspiral-plunge waveform modes are corrected by tidal terms
\begin{equation}
h_{\ell m}^{\textrm{insp-plunge}} = h_{\ell m}^{\textrm{insp-plunge, pm}} + h_{\ell m}^{\textrm{tidal}}\,,
\end{equation} 
where $h_{\ell m}^{\textrm{insp-plunge, pm}}$ is discussed in Section~II.B of Ref.~\cite{Bohe:2016gbl} and $h_{\ell m}^{\textrm{tidal}}$ is given by Eqs. (A14)-(A17) of Ref.~\cite{Damour:2012yf}. Following Ref.~\cite{Dietrich:2017feu}, in the computation of $h_{22}^{\textrm{tidal}}$ we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.

The inspiral-plunge orbital dynamics is obtained by numerically solving Hamilton's equations for the EOB Hamiltonian supplemented by a radiation-reaction force that is proportional to the GW flux derived from the $h_{\ell m}^{\textrm{insp-plunge}}$'s.

@@

--Tapering fit to NR.

--Universal relations to relate all matter parameters to quadrupolar $\Lambda$.

--Results in 5 waveform parameters $\vec x=\{q, S_1, S_2, \Lambda_1, \Lambda_2\}$.

>>>>

%% ________________________________________________________
\section{Aligned-spin, dynamical tides EOB model}
Ever since its inception~\cite{Buonanno:1998gg}, the effective-one-body (EOB) approach to the general-relativistic 2-body problem has proven successful in modeling the dynamics and GW emission of binaries of compact objects. State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match numerical-relativity (NR) simulations of spinning, nonprecessing binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to almost extremal for equal-mass) binaries. The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}. This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu}, which become especially relevant in the strong-field regime for 2-body systems that contain matter --- typically binary neutron stars (BNSs) or neutron-star/black hole binaries (NSBHs). The main effect of tides is to make the gravitational interaction more attractive with respect to the vacuum case. In particular, Refs.~\cite{Hinderer:2016eia,Steinhoff:2016rfi} built upon the aligned-spin BBH EOB model of Ref.~\cite{Taracchini:2013rva} and proposed a way to include the effect of dynamical tides. Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion --- be it a black hole or another neutron star. Because of binary motion, the forcing tidal field varies at the orbital frequency. Thus, in the late stages of the inspiral, the characteristic $f$-mode frequency of the neutron star can be dynamically approached, resulting in a resonant excitation of such mode. The net effect is an amplification of tidal effects as compared to the adiabatic limit, which assumes that the $f$-mode frequency is much larger than the frequency of the forcing tidal field. Dynamical tidal effects are implemented in the EOB model through a modification of the radial potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime.

In this paper we adopt the tidally-augmented expression of $\Delta_u$ that is discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}, namely
\begin{equation}
\Delta_u = \Delta_u^{\textrm{pm}} + \Delta_u^{\textrm{DT}}\,,
\end{equation}
where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass term of the underlying BBH EOB model of Ref.~\cite{Bohe:2016gbl} (see Eq.~(2.2) therein) and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides. We consider the general case of a binary that may consist of two neutron stars; if either component of the binary is a black hole, then vanishing values of the tidal polarizabilities are employed in the formulas below. Let $A$ and $B$ be the body labels. Let $M_{A,B}$ be the mass of either body and $M=M_A+M_B$ be the total mass of the binary.  Explicitly, including quadrupolar and octupolar tidal effects, we have
\begin{widetext}
\begin{align}
 \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)X_A^4 X_B\,u^6  \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
 &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,,
\end{align}
\end{widetext}
where $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities. Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies. In particular, the dimensionless dynamical tidal polarizability reads
\begin{equation}
\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})\,,
\end{equation}
where $\Lambda_{\ell}^{A,B}$ is the dimensionless adiabatic tidal polarizability
\begin{equation}
\Lambda_{\ell}^{A,B}=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\,,
\end{equation}
with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the neutron star compactness, which depends on the areal radius $R_{A,B}$ of the neutron star. Here $\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor given in Eq.~(11) of Ref.~\cite{Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole angular frequency, $\omega_{0\ell}^{A,B}$.

Tidal effects also enter the dissipative part of the model. In particular, the point-mass $\ell=2,3$ inspiral-plunge waveform modes are corrected by tidal terms
\begin{equation}
h_{\ell m}^{\textrm{insp-plunge}} = h_{\ell m}^{\textrm{insp-plunge, pm}} + h_{\ell m}^{\textrm{tidal}}\,,\label{hlminspplunge}
\end{equation} 
where $h_{\ell m}^{\textrm{insp-plunge, pm}}$ is discussed in Section~II.B of Ref.~\cite{Bohe:2016gbl} and $h_{\ell m}^{\textrm{tidal}}$ is given by Eqs. (A14)-(A17) of Ref.~\cite{Damour:2012yf}. Following Ref.~\cite{Dietrich:2017feu}, in the computation of $h_{22}^{\textrm{tidal}}$ we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.

The inspiral-plunge orbital dynamics is obtained by numerically integrating Hamilton's equations for the EOB Hamiltonian supplemented by a radiation-reaction force that is proportional to the GW flux derived from the $h_{\ell m}^{\textrm{insp-plunge}}$'s: see Eqs.~(5) and (6) in Ref.~\cite{Dietrich:2017feu}. The numerical integration starts from quasicircular initial conditions that are built following the prescription of Ref.~\cite{Buonanno:2005xu}. We exploit a fit for the (2,2)-mode peak frequency of NR BNS simulations~\cite{Bernuzzi:2015rla} to estimate the final orbital frequency of the EOB inspiral-plunge dynamics. When such frequency is being approached (to within some tolerance), we stop the numerical evolution if a peak in the orbital frequency is detected or if the binary stops to monotonically plunge, similarly to what is done in the BBH case.

Let us now discuss the construction of the full waveform, from the inspiral to the post-merger stage. In this work we only consider the dominant mode, $(\ell,m)=(2,2)$. The inspiral-plunge GW emission is obtained by evaluating Eq.~(\ref{hlminspplunge}) along the orbital dynamics up to $t=t_{\textrm{match}}^{22}$, that is the time when either the amplitude or the GW frequency of $h_{22}^{\textrm{insp-plunge}}$ reaches a global maximum, whichever occurs first. We model the post-merger GW emission --- i.e., at $t>t_{\textrm{match}}^{22}$ --- by a signal that tapers down in amplitude and has a constant GW frequency. In particular, we employ the amplitude of the merger-ringdown of the BBH model (as given by the magnitude of Eq.~(2.10) of Ref.~\cite{Bohe:2016gbl}), using as inputs the mass ratio and spins of the BNS/NSBH of interest. This choice is in reasonable qualitative agreement with the post-merger behavior of the two equal-mass nonspinning NR BNS simulations that were analyzed in Ref.~\cite{Hinderer:2016eia}. The constant value of the GW frequency is taken to be that of $h_{22}^{\textrm{insp-plunge}}$ at time $t=t_{\textrm{match}}^{22}$. A more sophisticated, NR-informed, time-domain model of the post-merger emission will be part of future investigations.

As described above, an EOB waveform with dynamical tides depends on 11 intrinsic parameters: the mass ratio $0\leq q=M_B/M_A\leq1$, the dimensionless spin components along the angular momentum of the binary $-1 \leq \chi_{A,B} \leq1$, the adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, the $\ell=2,3$ $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$. We note that this list does not include the total mass $M$ of the system, since the point-mass part of the model is scale-invariant and tidal corrections only depend on $X_A=1/(1+q)$, $X_B=q/(1+q)$, and $\Lambda_2^{A,B}$, which is inherent to the PN analysis of tidal effects~\cite{Flanagan:2007ix}. To reduce the dimensionality of the intrinsic parameter space, we exploit universal (i.e., equation-of-state-independent) relations that are available in the literature. In particular, Eq.~(60) of Ref.~\cite{Yagi:2013sva} with the coefficients from the first row of Table I therein expresses $\Lambda_3^{A,B}$ as a function of $\Lambda_2^{A,B}$ with an accuracy to within about 10\%, depending on the equation of state and the value of $\Lambda_2^{A,B}$. Moreover, Eq. (3.5) of Ref.~\cite{Chan:2014kua} with the coefficients from the first column of Table I therein gives $\omega_{02}^{A,B}$ as a function of $\Lambda_2^{A,B}$ to within a few percent, while the same equation but with the coefficients from the second column of Table I therein gives $\omega_{03}^{A,B}$ as a function of $\Lambda_3^{A,B}$ to within a few percent. Thanks to these relations, the waveform only depends on 5 intrinsic parameters, which we denote as $\boldsymbol{x}=\{q, \chi_A,\chi_B, \Lambda_2^A,\Lambda_2^B\}$.

@@
